---
title: 'Analysis Report (`r params$condition` vs. `r params$control`)'
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    df_print: paged

params:
  pval_cutoff: 0.05
  lfc_cutoff: 0.58
  low_counts_cutoff: 10
  meta_data: ''
  counts_data: ''
  condition: ''
  control: ''
  conversion_df: ''
  mm_h: ''
  mm_c5_bp: ''
  mm_c5_cc: ''
  mm_c5_mf: ''
---

<style>
pre code, pre, code {
  white-space: pre !important;
  overflow-x: auto !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

```{r setup, include = F}
options(width = 10000)

library(tidyverse)
library(scales)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(fgsea)
library(VennDiagram)

select <- dplyr::select

pval_cutoff <- params$pval_cutoff
lfc_cutoff <- params$lfc_cutoff
low_counts_cutoff <- params$low_counts_cutoff

meta_data <- params$meta_data
conversion_df <- params$conversion_df
mm_h <- params$mm_h
mm_c5_bp <- params$mm_c5_bp
mm_c5_cc <- params$mm_c5_cc
mm_c5_mf <- params$mm_c5_mf

condition <- params$condition
control <- params$control
counts_data <- params$counts_data

counts_data <- data.frame(counts_data %>% select((meta_data %>% filter(experimental_class_type %in% c(condition, control), `Assay Type` == 'RNA-Seq'))$Run), row.names = counts_data$Geneid)

colors_default <- c('#00d5dc', '#ff877e')


get_DESeqDataSet_obj <- function(cnts, design_formula) {
  
  col_data <- meta_data %>%
    filter(Run %in% names(cnts)) %>%
    select(Run, experimental_class_type, regime, treatment)
  
  col_data <- col_data[match(names(cnts), col_data$Run), ] %>%
    column_to_rownames(var = 'Run')
  
  print(all(rownames(col_data) %in% colnames(cnts)))
  print(all(rownames(col_data) == colnames(cnts)))
  
  dds <- DESeqDataSetFromMatrix(countData = as.matrix(round(cnts)),
                                colData = col_data,
                                design = design_formula)
  
  print(summary(dds))
  
  # Filter rows with low counts
  keep <- rowSums(counts(dds)) >= low_counts_cutoff
  dds <- dds[keep, ]
  
  print(summary(dds))
  
  dds
}

get_transformed_data <- function(dds, transformation_type = 'rlog', blind = T) {
  if (transformation_type == 'vst') {
    dds_transformed <- vst(dds, blind = blind)
  } else {
    dds_transformed <- rlog(dds, blind = blind)
  }
  
  list(
    matrix = assay(dds_transformed),
    pca = prcomp(t(assay(dds_transformed)))
  )
}

get_fgsea_res <- function(rank_vec, genesets_list) {
  fgsea(
    pathways = genesets_list,
    stats = rank_vec,
    nperm = 1000,
    minSize = 15,
    maxSize = 500
  ) %>%
    as_tibble() %>%
    arrange(desc(NES))
}

plot_enrichment_table <- function(gsea_res, rank_vec, genesets_list, n_top = 10) {
  top_pathways_up <- gsea_res %>% filter(ES > 0) %>% arrange(padj) %>% head(n_top)
  top_pathways_down <- gsea_res %>% filter(ES < 0) %>% arrange(padj) %>% head(n_top)
  top_pathways <- bind_rows(top_pathways_up, top_pathways_down) %>% arrange(-NES)

  plotGseaTable(
    genesets_list[top_pathways$pathway], 
    rank_vec, 
    gsea_res, 
    gseaParam = 0.5
  )
}
```

```{r, echo = F}
writeLines(str_c('pval_cutoff: ', pval_cutoff))
writeLines(str_c('lfc_cutoff: ', lfc_cutoff))
writeLines(str_c('low_counts_cutoff: ', low_counts_cutoff))
```


# General statistics

```{r}
# Number of samples
length(counts_data)

# Number of genes
nrow(counts_data)

# Total counts
colSums(counts_data)
```

```{r, echo = F, fig.width = 10}
counts_df <- data.frame(count = colSums(counts_data)) %>% 
  rownames_to_column(var = 'Run') %>% 
  as_tibble() %>% 
  left_join(meta_data %>% select(Run, experimental_class_type, regime), by = 'Run')

counts_df %>% ggplot(aes(x = Run, y = count, fill = regime)) +
  geom_col() +
  scale_y_continuous(labels = label_number(suffix = 'M', scale = 1e-6)) +
  scale_fill_manual(values = colors_default) +
  xlab('') + ylab('') +
  theme(text = element_text(size = 9))
```


# Create DDS objects

```{r, message = F}
# Create DESeqDataSet object
dds <- get_DESeqDataSet_obj(counts_data, ~ experimental_class_type)
colData(dds)
```


# Sample-to-sample comparisons

```{r}
# Transform data (blinded rlog)
rld <- get_transformed_data(dds)
```

## PCA plot

```{r}
pca <- rld$pca
pca_df <- cbind(as.data.frame(colData(dds)) %>% rownames_to_column(var = 'name'), pca$x)
summary(pca)

ggplot(pca_df, aes(x = PC1, y = PC2, color = regime)) +
  geom_point() +
  geom_text(aes(label = name), position = position_nudge(y = -2), show.legend = F, size = 3) +
  scale_color_manual(values = colors_default) +
  scale_x_continuous(expand = c(0.2, 0))
```

## Correlation heatmap

```{r}
pheatmap(
  cor(rld$matrix),
  annotation_col = as.data.frame(colData(dds)) %>% select(regime),
  color = brewer.pal(8, 'YlOrRd')
)
```


# Wald test results

```{r, message = F}
# DE analysis using Wald test
dds_full <- DESeq(dds)
colData(dds_full)

# Wald test results
res <- results(
  dds_full,
  contrast = c('experimental_class_type', condition, control),
  alpha = pval_cutoff
)
res

mcols(res)
summary(res)
```

## Summary details

```{r, echo = F}
res_df <- res %>%  # dataframe version
  data.frame() %>% 
  rownames_to_column(var = 'ensembl_gene_id') %>% 
  as_tibble()

res_sig_df <- res_df %>%  # filter only DE genes
  filter(padj < pval_cutoff) %>% 
  arrange(padj)
```

```{r}
# Upregulated genes (LFC > 0)
res_sig_df %>% filter(log2FoldChange > 0)

# Downregulated genes (LFC < 0)
res_sig_df %>% filter(log2FoldChange < 0)

# Outliers (pvalue and padj are NA)
res[which(is.na(res$pvalue)), ]

# Low counts (only padj is NA)
res[which(is.na(res$padj) & !is.na(res$pvalue)), ]
```


# Shrunken LFC results

```{r, message = F}
plotMA(res)

# Shrunken LFC results
res_shrunken <- lfcShrink(
  dds_full,
  coef = str_c('experimental_class_type_', condition, '_vs_', control),
  type = 'apeglm'
)
res_shrunken

plotMA(res_shrunken)

mcols(res_shrunken)
summary(res_shrunken, alpha = pval_cutoff)
```

## Summary details

```{r, echo = F}
res_shrunken_df <- res_shrunken %>%  # dataframe version
  data.frame() %>% 
  rownames_to_column(var = 'ensembl_gene_id') %>% 
  as_tibble()

res_shrunken_sig_df <- res_shrunken_df %>%  # filter only DE genes
  filter(padj < pval_cutoff) %>% 
  arrange(padj)
```

```{r}
# Upregulated genes (LFC > 0)
res_shrunken_sig_df %>% filter(log2FoldChange > 0)

# Downregulated genes (LFC < 0)
res_shrunken_sig_df %>% filter(log2FoldChange < 0)

# Outliers (pvalue and padj are NA)
res_shrunken[which(is.na(res_shrunken$pvalue)), ]

# Low counts (only padj is NA)
res_shrunken[which(is.na(res_shrunken$padj) & !is.na(res_shrunken$pvalue)), ]
```


# Visualizing results

```{r, include = F}
counts_sig_norm <- counts(dds_full, normalized = T) %>%  # normalized counts
  data.frame() %>%
  rownames_to_column(var = 'ensembl_gene_id') %>% 
  as_tibble() %>% 
  filter(ensembl_gene_id %in% res_sig_df$ensembl_gene_id)

counts_sig_log <- rld$matrix %>%  # log-transformed counts
  data.frame() %>%
  rownames_to_column(var = 'ensembl_gene_id') %>% 
  as_tibble() %>% 
  filter(ensembl_gene_id %in% res_sig_df$ensembl_gene_id)
```

## Heatmaps

```{r}
# Plot normalized counts (z-scores)
pheatmap(counts_sig_norm[2:7], 
         color = brewer.pal(8, 'YlOrRd'), 
         cluster_rows = T, 
         show_rownames = F,
         annotation_col = as.data.frame(colData(dds)) %>% select(regime),
         border_color = NA,
         fontsize = 10,
         scale = 'row',
         fontsize_row = 10, 
         height = 20)

# Plot log-transformed counts
pheatmap(counts_sig_log[2:7], 
         color = rev(brewer.pal(8, 'RdYlBu')), 
         cluster_rows = T, 
         show_rownames = F,
         annotation_col = as.data.frame(colData(dds)) %>% select(regime),
         border_color = NA,
         fontsize = 10,
         fontsize_row = 10, 
         height = 20)

# Plot log-transformed counts (top 24 DE genes)
pheatmap((counts_sig_log %>% filter(ensembl_gene_id %in% res_sig_df$ensembl_gene_id[1:24]))[2:7],
         color = rev(brewer.pal(8, 'RdYlBu')), 
         cluster_rows = T, 
         show_rownames = F,
         annotation_col = as.data.frame(colData(dds)) %>% select(regime), 
         fontsize = 10,
         fontsize_row = 10, 
         height = 20)
```

## Volcano plots

```{r, warning = F}
# Unshrunken LFC
res_df %>% 
  mutate(
    sig_threshold = if_else(
      padj < pval_cutoff & abs(log2FoldChange) >= lfc_cutoff,
      if_else(log2FoldChange > 0, 'DE-up', 'DE-down'),
      'non-DE'
    )
  ) %>% 
  filter(!is.na(sig_threshold)) %>% 
  ggplot() +
  geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = sig_threshold)) +
  scale_color_manual(values = c('blue', 'red', 'gray')) +
  xlab('log2 fold change') + 
  ylab('-log10 adjusted p-value')

# Shrunken LFC
res_shrunken_df %>% 
  mutate(
    sig_threshold = if_else(
      padj < pval_cutoff & abs(log2FoldChange) >= lfc_cutoff,
      if_else(log2FoldChange > 0, 'DE-up', 'DE-down'),
      'non-DE'
    )
  ) %>% 
  filter(!is.na(sig_threshold)) %>% 
  ggplot() +
  geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = sig_threshold)) +
  scale_color_manual(values = c('blue', 'red', 'gray')) +
  xlab('log2 fold change') + 
  ylab('-log10 adjusted p-value')
```


# GSEA (all)

```{r, echo = F, warning = F, message = F}
# Shrunken LFC
rank_lfc_df <- res_shrunken_df %>%
  left_join(conversion_df, by = 'ensembl_gene_id') %>% 
  filter(!is.na(entrezgene_id)) %>% 
  arrange(desc(abs(log2FoldChange))) %>% 
  group_by(entrezgene_id) %>% 
  summarise(rank = dplyr::first(log2FoldChange)) %>% 
  filter(!is.na(rank))

rank_lfc <- rank_lfc_df$rank
names(rank_lfc) <- rank_lfc_df$entrezgene_id

# Wald stat
rank_stat_df <- res_df %>% 
  left_join(conversion_df, by = 'ensembl_gene_id') %>%  
  filter(!is.na(entrezgene_id)) %>% 
  group_by(entrezgene_id) %>% 
  summarise(rank = base::mean(stat, na.rm = T)) %>% 
  filter(!is.na(rank))

rank_stat <- rank_stat_df$rank
names(rank_stat) <- rank_stat_df$entrezgene_id

# Rank: sign(LFC) * -log10(pvalue)
rank_pval_df <- res_df %>% 
  left_join(conversion_df, by = 'ensembl_gene_id') %>% 
  filter(!is.na(entrezgene_id)) %>% 
  mutate(rank = sign(log2FoldChange) * -log10(pvalue)) %>% 
  group_by(entrezgene_id) %>% 
  summarise(rank = base::mean(rank, na.rm = T)) %>% 
  filter(!is.na(rank))

rank_pval <- rank_pval_df$rank
names(rank_pval) <- rank_pval_df$entrezgene_id
```

## Hallmark genesets

```{r, warning = F, fig.width = 13}
# Shrunken LFC
get_fgsea_res(rank_lfc, mm_h) %>% plot_enrichment_table(rank_lfc, mm_h)
```

```{r, warning = F, fig.width = 13}
# Wald stat
get_fgsea_res(rank_stat, mm_h) %>% plot_enrichment_table(rank_stat, mm_h)
```

```{r, warning = F, fig.width = 13}
# Rank: sign(LFC) * -log10(pvalue)
get_fgsea_res(rank_pval, mm_h) %>% plot_enrichment_table(rank_pval, mm_h)
```

## GO biological process

```{r, warning = F, fig.width = 13}
# Shrunken LFC
get_fgsea_res(rank_lfc, mm_c5_bp) %>% plot_enrichment_table(rank_lfc, mm_c5_bp)
```

```{r, warning = F, fig.width = 13}
# Wald stat
get_fgsea_res(rank_stat, mm_c5_bp) %>% plot_enrichment_table(rank_stat, mm_c5_bp)
```

```{r, warning = F, fig.width = 13}
# Rank: sign(LFC) * -log10(pvalue)
get_fgsea_res(rank_pval, mm_c5_bp) %>% plot_enrichment_table(rank_pval, mm_c5_bp)
```

## GO cellular component

```{r, warning = F, fig.width = 13}
# Shrunken LFC
get_fgsea_res(rank_lfc, mm_c5_cc) %>% plot_enrichment_table(rank_lfc, mm_c5_cc)
```

```{r, warning = F, fig.width = 13}
# Wald stat
get_fgsea_res(rank_stat, mm_c5_cc) %>% plot_enrichment_table(rank_stat, mm_c5_cc)
```

```{r, warning = F, fig.width = 13}
# Rank: sign(LFC) * -log10(pvalue)
get_fgsea_res(rank_pval, mm_c5_cc) %>% plot_enrichment_table(rank_pval, mm_c5_cc)
```

## GO molecular function

```{r, warning = F, fig.width = 13}
# Shrunken LFC
get_fgsea_res(rank_lfc, mm_c5_mf) %>% plot_enrichment_table(rank_lfc, mm_c5_mf)
```

```{r, warning = F, fig.width = 13}
# Wald stat
get_fgsea_res(rank_stat, mm_c5_mf) %>% plot_enrichment_table(rank_stat, mm_c5_mf)
```

```{r, warning = F, fig.width = 13}
# Rank: sign(LFC) * -log10(pvalue)
get_fgsea_res(rank_pval, mm_c5_mf) %>% plot_enrichment_table(rank_pval, mm_c5_mf)
```


# GSEA (DE)

```{r, echo = F, warning = F, message = F}
# Shrunken LFC
rank_lfc_df <- res_shrunken_df %>%
  filter(padj < pval_cutoff) %>% 
  left_join(conversion_df, by = 'ensembl_gene_id') %>% 
  filter(!is.na(entrezgene_id)) %>% 
  arrange(desc(abs(log2FoldChange))) %>% 
  group_by(entrezgene_id) %>% 
  summarise(rank = dplyr::first(log2FoldChange)) %>% 
  filter(!is.na(rank))

rank_lfc <- rank_lfc_df$rank
names(rank_lfc) <- rank_lfc_df$entrezgene_id

# Wald stat
rank_stat_df <- res_df %>% 
  filter(padj < pval_cutoff) %>% 
  left_join(conversion_df, by = 'ensembl_gene_id') %>%  
  filter(!is.na(entrezgene_id)) %>% 
  group_by(entrezgene_id) %>% 
  summarise(rank = base::mean(stat, na.rm = T)) %>% 
  filter(!is.na(rank))

rank_stat <- rank_stat_df$rank
names(rank_stat) <- rank_stat_df$entrezgene_id

# Rank: sign(LFC) * -log10(pvalue)
rank_pval_df <- res_df %>% 
  filter(padj < pval_cutoff) %>% 
  left_join(conversion_df, by = 'ensembl_gene_id') %>% 
  filter(!is.na(entrezgene_id)) %>% 
  mutate(rank = sign(log2FoldChange) * -log10(pvalue)) %>% 
  group_by(entrezgene_id) %>% 
  summarise(rank = base::mean(rank, na.rm = T)) %>% 
  filter(!is.na(rank))

rank_pval <- rank_pval_df$rank
names(rank_pval) <- rank_pval_df$entrezgene_id
```

## Hallmark genesets

```{r, warning = F, fig.width = 13}
# Shrunken LFC
get_fgsea_res(rank_lfc, mm_h) %>% plot_enrichment_table(rank_lfc, mm_h)
```

```{r, warning = F, fig.width = 13}
# Wald stat
get_fgsea_res(rank_stat, mm_h) %>% plot_enrichment_table(rank_stat, mm_h)
```

```{r, warning = F, fig.width = 13}
# Rank: sign(LFC) * -log10(pvalue)
get_fgsea_res(rank_pval, mm_h) %>% plot_enrichment_table(rank_pval, mm_h)
```

## GO biological process

```{r, warning = F, fig.width = 13}
# Shrunken LFC
get_fgsea_res(rank_lfc, mm_c5_bp) %>% plot_enrichment_table(rank_lfc, mm_c5_bp)
```

```{r, warning = F, fig.width = 13}
# Wald stat
get_fgsea_res(rank_stat, mm_c5_bp) %>% plot_enrichment_table(rank_stat, mm_c5_bp)
```

```{r, warning = F, fig.width = 13}
# Rank: sign(LFC) * -log10(pvalue)
get_fgsea_res(rank_pval, mm_c5_bp) %>% plot_enrichment_table(rank_pval, mm_c5_bp)
```

## GO cellular component

```{r, warning = F, fig.width = 13}
# Shrunken LFC
get_fgsea_res(rank_lfc, mm_c5_cc) %>% plot_enrichment_table(rank_lfc, mm_c5_cc)
```

```{r, warning = F, fig.width = 13}
# Wald stat
get_fgsea_res(rank_stat, mm_c5_cc) %>% plot_enrichment_table(rank_stat, mm_c5_cc)
```

```{r, warning = F, fig.width = 13}
# Rank: sign(LFC) * -log10(pvalue)
get_fgsea_res(rank_pval, mm_c5_cc) %>% plot_enrichment_table(rank_pval, mm_c5_cc)
```

## GO molecular function

```{r, warning = F, fig.width = 13}
# Shrunken LFC
get_fgsea_res(rank_lfc, mm_c5_mf) %>% plot_enrichment_table(rank_lfc, mm_c5_mf)
```

```{r, warning = F, fig.width = 13}
# Wald stat
get_fgsea_res(rank_stat, mm_c5_mf) %>% plot_enrichment_table(rank_stat, mm_c5_mf)
```

```{r, warning = F, fig.width = 13}
# Rank: sign(LFC) * -log10(pvalue)
get_fgsea_res(rank_pval, mm_c5_mf) %>% plot_enrichment_table(rank_pval, mm_c5_mf)
```


# System Info

```{r}
sessionInfo()
```
